Evaluation of a Cache-Oblivious Data Structure [draft-1]

### 2008-05-31 paper/paper.tex
Updates to the draft paper.

One of the reasons this assumption fails is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (fast\-er) main memory when the latter is in short supply. Another reason is that as processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations. Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem, but this increases the differences in access time between different types of memory.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l l l l }
\hline
\textbf{Type} & \textbf{Total Size} & \textbf{Block Size} & \textbf{Access Time} \\
\hline
Registers  &   32 bytes & 4 bytes & 1 cycle \\
L1 Cache   &   32 KB    & 64 bytes & 3 cycles \\
L2 Cache   &    1 MB    & 64 bytes & 9 cycles \\
RAM        &   ~2 GB    & 1 KB     & 50-100 cycles \\
Disk       & ~300 GB    & 4 KB     & 5,000-10,000 cycles \\
\hline
\end{tabular}
\caption{Sizes and access times of various types of storage}
\label{tab-memhier}
\end{center}
\end{table}

@@

Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice. Although some research has been done on cache-oblivious data structures, they are seldom used in practice. The main contribution of this paper is to give some insight in the practical merit of cache-oblivious data structures.

@@

\subsection{Cache-Oblivious Memory Model}
A different approach to generalizing the external memory model was taken by Prokop \cite{prokop1999coa} who proposed the cache-oblivious model, which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory. Data is transferred between the two in aligned data blocks of size B. In contrast with the hierarchical memory model, the application does not have explicit control over the transferring of blocks between the two memories. Instead, it is assumed that an optimal cache manager exists which minimizes the number of block transfers over the execution the program. Additionally (and in contrast with the external memory model) the values of parameters like M and B are not known to the application, so they cannot be used explicitly when defining algorithms and data structures. Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).

@@

Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion. This data structure is mostly of theoretic interest, as it is fairly complex to implement and incurs a relatively large overhead in practice.

Later, Bender, Duan, Iacono and Wu simplified the data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding support for finger searches. The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static tree in van Emde Boas layout (named for similarity to a different data structure proposed by van Emde Boas et al. \cite{vanemdeboas1976dai}).

@@

\section{Methods}
In order to gather empirical data, the research approach must be made more concrete. We will need to limit ourselves to a specific class of data structures, since algorithms and data structures offering different functionality cannot be compared in a meaningful way. Furthermore, we will need to select a proper scenario in which the data structures are evaluated, as conclusions on the practical merit of the data structures depend on the degree to which the test scenario is realistic.

Furthermore, we will need to define what practical performance is. We will measure two properties: primarily, the total running time of our test application, and secondarily the memory used. The rationale for selecting these metrics is that if the data structures perform identical functions and plenty of memory is available, the only observable difference in running a program using different data structures will be the run time. Memory use is of secondary interest because in practice memory may be limited, which would preclude the use of certain data structures that require a lot of memory to function.

@@

Depending on how states are inserted and extracted from the queue, the search
may be breadth-first (meaning that all states that are reachable in $N$ steps
from the initial state are expanded before any states that require more than
$N$ steps) or depth-first (meaning that states that are generated most recently
by the successor function are expanded first). We will use breadth-first
search, which means the queue operates by a first-in first-out principle.

@@

\subsection{Experimental Framework}
For our experimental framework, we need a collection of formal models that are
representative of those typically used for formal verification, and a way to
execute them. For the first part, we have looked at Spin \cite{spin}, a widely
used model checking tool. Spin uses a custom modeling language (called
PROMELA) to specify models and is distributed with a collection of example
models that are suitable for our experiments.

For the execution of these models we will use the NIPS VM \cite{weber2007evm}, a
high-performance virtual machine for state space generation that is easily
embeddable in our framework. Although the NIPS VM only executes bytecode in a
custom format, a PROMELA compiler is also available to generate the required
bytecode from the Spin models.\cite{nipsvm} The NIPS VM is preferred to the
Spin tools because it was designed to be embedded in other programs and as
such is more easily integrated in our framework.

@@

@@

When several values are inserted, some values may map to the same slot,
which is problematic if each slot can only store one value. There
are a lot of different ways to resolve this collision problem; we use
\emph{separate chaining} which means that slots do not store values directly,
but instead each slot stores a pointer to a singly-linked list of values that
map to that slot. This particular implementation technique is well-suited to
the scenario where values may have different sizes (as the slots only need to
store a fixed-size pointer and not a variable-size value) and maintains
relatively good performance when the number of values stored exceeds the size
of the index array \cite{sedgewick1998ac}.

@@

To insert a value, simply find the leaf page where it belongs. If this page
does not have enough free space to insert it, the page is split into two pages;
the median value from the old page is moved to the parent page, and two new pages
(containing the values less than respectively greater than the median) are
created. When the top-most page needs to be split, a new (empty) top-level page is
created and the height of the B-tree is increased by one. As a result, all leaf
nodes in a B-tree are at the same depth, and all pages stored are at least half
full, except possibly the root page.

% TODO: picture and description

B-trees can easily support the insertion of variable-length values, as long
as individual values fit in the size of a single page. However, there is no
obvious way to store values that are larger than the size of a single page.
Production-quality implementations handle this case separately. Our our
implementation does not support this requires stored values to
be smaller than the page size. Additionally, deletion of items has not been
implemented.

\subsubsection{Bender Set Implementation}
The first implementation challenge for the Bender set is that the description
given by Bender et al. assumes that all values stored in the set are of a
fixed size, which is not the case in our experimental framework. To work
around this, we create several separate instances of the Bender set with
different value sizes which are powers of two. When a value is to be inserted,
its size is rounded up to a power of two and it is inserted in the
corresponding set instance. This ensures that the amount of spaces wasted
remains below 50\% while still allowing values of various sizes to be inserted.
In the following description, we will assume that values are of a fixed size.

@@

According to this description (which follows the paper by Bender et al) a window
is rebalanced every time an element is inserted. As an optimization, our
implementation does not always rebalance a window. When there is free space
between the successor and predecessor of the value to be inserted, the value
is simply inserted in the middle. In our experiments, this optimization
yielded an reduction in execution time.

@@

Finally, the Bender set uses a complete binary tree as an index data structure
to efficiently find the successor element for a given value. The tree is
stored in memory in van Emde Boas lay-out (as described by Prokop) to allow
efficient cache-oblivious traversal.

@@

Therefore, we decided to measure wall clock time. To deal with variations
across multiple executions, each experiment is run three times, and the median
value is used for further analysis.

@@

In the results below, all values are reported relative to the values obtained
using the mock set implementation, which means the overhead of the test
framework is removed from the results.

\subsubsection{System Configuration}
The experiments where performed on a 64-bit Linux (kernel 2.6) system
with an Intel Xeon E5320 processor (1.86 GHz, 4MB cache) and 8GB of main
memory and 2GB swap file. Although the system is a multiprocessor system
(and the E5320 is a dual-core processor) all code is single-threaded so
only a single core is used when executing tests.

% Describe data set used; per model:
%   average state size?
%   total size?
%   ratio of insertions/queries?

@@

\begin{table}
\begin{center}
\begin{tabular}{ l l l l }
\hline
\textbf{Model} & \textbf{Parameters} & \textbf{Iterations} & \textbf{Transitions} \\
\hline
Eratosthenes & $N=40$ &  1,019,960 &  4,923,218 \\
Leader2      & $N=5$  &  5,950,945 & 23,856,363 \\
Peterson\_N  & $N=4$  & 10,000,000 & 37,434,411 \\
\hline
\end{tabular}
\caption{Properties of test cases used}
\label{tab-cases}
\end{center}
\end{table}

@@

Figure \ref{fig-eratosthenes-hash_wctime} shows the execution times for the hash
table. About a million states need to be stored and as expected, the hash table
with an index size that is close to this number performs best. The hash table
with a small index (100,000 slots) performs well initially, but gets slower as
it becomes too full, at which point each slot in the hash table has to store a
long list of values that map to that slot.
The hash table with a large index (10,000,000 slots) is a bit slower than the hash
table with an index size of 1,000,000, possible because of cache misses caused by
the larger index size.
As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
usage by the different hash tables is in the constant size of the index.
Despite not being the fastest in this case, we will use the hash table with an
index size of 10,000,000 for further experiments, as other test cases require
storing well over 1 million states and we want to use a single size that works
well for all test cases.

Figure \ref{fig-eratosthenes-btree_wctime} and Figure
\ref{fig-eratosthenes-btree_vss} show the execution time and memory usage
for the B-tree with page sizes ranging from 1KB to 16KB. The execution platform
maps memory in 4KB pages, which would suggest that using pages less than 4KB
makes little sense. Indeed, the B-tree with 1KB pages seems to perform worst,
but if we take the memory use into account, this is more likely due to the fact
that fewer items fit on a single page (which means a relatively large portion of
the page remains unused) than anyting else.
The 4KB and 16KB pages are relatively close (both in execution time and memory
usage); we select the 16KB page size for further experiments because state sizes
are larger in other test cases, in which case the 4KB page size may cause similar
problems as the 1KB page size here.

@@

% TODO: test with larger density!

Now that we have established the parameters to use for the other test cases.
The execution times for the three testcases are presented in figures
\ref{fig-eratosthenes_wctime}, \ref{fig-leader2_wctime} and
\ref{fig-peterson_wctime}, while the corresponding memory usage is presented in
figures \ref{fig-eratosthenes_vss}, \ref{fig-leader2_vss} and \ref{fig-peterson_vss} respectively.

@@

@@

@@

@@

% What else to include:
%   queue size over iterations
%       for each program
%
%   insertions per query:
%       for each program
%

>>>>

The main cause for this is the increasing difference between processor and memory speeds. As processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations. Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem.
Similar developments have taken place on the boundary between main memory and disk-based storage.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l l l l }
\hline
\textbf{Type} & \textbf{Total Size} & \textbf{Block Size} & \textbf{Access Time} \\
\hline
Registers  &   32~bytes & 4~bytes & 1 cycle \\
L1 Cache   &   32~KB    & 64~bytes & 3 cycles \\
L2 Cache   &    1~MB    & 64~bytes & 9 cycles \\
RAM        &  $\pm$ 2 GB    & 1~KB     & 50-100 cycles \\
Disk       &  $\pm$ 300 GB    & 4~KB     & 5,000-10,000 cycles \\
\hline
\end{tabular}
\caption{Sizes and access times of various types of storage}
\label{tab-memhier}
\end{center}
\end{table}

@@

Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice. Although some research has been done on cache-oblivious data structures, at the time of writing there is no evidence of their use in practice. The main goal for this paper is to give some insight in the practical merit of cache-oblivious data structures.

In the following pages we will give an overview of the different available memory models and explain why the cache-oblivious model is of particular interest. We will then describe what our goals for this research project were and how our work relates to previous research.
A large part of the paper will be dedicated to a description of our research methodology. Finally, we will present and discuss our results, and draw a conclusion on the applicability of cache-oblivious data structures.

\section{Previous Work}
The cache-oblivious memory model is not the only or the first model that was developed as a more realistic alternative to the uniform memory model. We will briefly describe some of these alternatives here.

@@

\subsection{Cache-Oblivious Memory Model}
A different approach to generalizing the external memory model was taken by Prokop \cite{prokop1999coa} who proposed the cache-oblivious model.
In this model, there is an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory.
Data is transferred between the two in aligned data blocks of size B.
In contrast with the hierarchical memory model, the application does not have explicit control over the transferring of blocks between the two memories.
Instead, it is assumed that an optimal cache manager exists which minimizes the number of block transfers over the execution the program.
Additionally (and in contrast with the external memory model) the values of parameters like M and B are not known to the application, so they cannot be used explicitly when defining algorithms and data structures.
Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).

@@

Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
Later, Bender, Duan, Iacono and Wu simplified this data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding support for additional operations (finger searches in particular).
This data structure will be explained in detail in Section \ref{sect-bender-set}.
The authors note that a chief advantage of their data structure over the previously described on is that it is less complex, more easily implementable and therefore more suitable for practical use.

@@

From the available publications we can conclude that only a minority of the research on the cache-oblivious memory model compares the practical performance of newly proposed data structures with that of established data structures.
Contrary to what theoretical analysis suggests, the practical results that are available so far fail to show the superior performance of cache-oblivious data structures.
Therefore, additional research is needed to determine more precisely to which extent cache-oblivious data structures are useful as a building block for practical work; this paper will provide some insight in this regard.

\section{Research Methods}
In order to gather empirical data, the research approach must be made more concrete. We will need to limit ourselves to a specific class of data structures, since algorithms and data structures offering different functionality cannot be compared in a meaningful way. Furthermore, we will need to select a proper scenario in which the data structures are evaluated, as conclusions on the practical merit of the data structures depend on the degree to which the test scenario is realistic.

We also need to define more accurately what we mean by practical performance. We will execute our test application and measure two properties: primarily, the total running time, and secondarily the memory used. The rationale for selecting these metrics is that if the data structures perform identical functions and plenty of memory is available, the only observable difference in running a program using different data structures will be the run time. Memory use is of secondary interest because in practice memory may be limited, which would preclude the use of certain data structures that require a lot of memory to function.

@@

Although variations are possible, our queue operates by a first-in, first-out
principle, meaning that meaning that all states that are reachable in $N$ steps
from the initial state are expanded before any states that require more than
$N$ steps.

@@

\subsection{Experimental Framework}
For our experimental framework, we need a collection of formal models that are
representative of those typically used for formal verification, and a way to
execute them. For the first part, we have looked at Spin \cite{holzmann2004smc},
a widely used model checking tool.
Spin uses a custom modeling language (called PROMELA) to specify models and is
distributed with a collection of example models that are suitable for our
experiments.

For the execution of these models we will use the NIPS VM \cite{weber2007evm}, a
high-performance virtual machine for state space generation that is easily
embeddable in our framework. Although the NIPS VM only executes bytecode in a
custom format, a PROMELA compiler is also available to generate the required
bytecode from the Spin models \cite{nipsvm}. The NIPS VM is preferred to the
Spin tools because it was designed to be embedded in other programs and as
such is more easily integrated in our framework.

@@

It should be noted that since we only need a limited subset of the functionality
offered by the set data structures for our test application, we did not
implement any operations that were not required to perform our experiments.
However, we did not change the design of the data structures to take advantage
of this. That means that any additional operations could be implemented without
structural changes to our existing implementation.

@@

When several values are inserted, some values may map to the same slot,
which is problematic if each slot can only store one value. There
are a lot of different ways to resolve this collision problem; we use
\emph{separate chaining}, which means that slots do not store values directly,
but instead each slot stores a pointer to a singly-linked list of values that
map to that slot. This particular implementation technique is well-suited to
the scenario where values may have different sizes (as the slots only need to
store a fixed-size pointer and not a variable-size value) and maintains
relatively good performance when the number of values stored exceeds the size
of the index array \cite{sedgewick1998ac}.

@@

New values are inserted into a leaf page which can be determined by traversing
the tree.
If this leaf page does not have enough free space to insert the new value,
the page will have to be \emph{split}: the median value is selected, and the old
page is replaced by two new pages containing the values less than respectively
greater than this median value, while the median value itself is moved to the
parent page.
When the top-most page needs to be split, a new (empty) top-level page is
created and the height of the B-tree is increased by one. As a result, all leaf
nodes in a B-tree are at the same depth, and all pages stored are at least half
full, except possibly the root page.

B-trees can easily support the insertion of variable-length values, as long
as each individual value fits in a single page. However, values with
a size larger than the size of a single page must be handled separately.
Our implementation does not support this and therefore requires all
stored values to be smaller than the pages.

\subsubsection{Bender Set Implementation}
\label{sect-bender-set}
The first implementation challenge for the Bender set is that the description
given by Bender et al. assumes that all values stored in the set are of a
fixed size, which is not the case in our experimental framework. To work
around this, we create several separate instances of the Bender set with
different value sizes which are powers of two. When a value is to be inserted,
its size is rounded up to a power of two and it is inserted in the
corresponding set instance. This ensures that the amount of spaces wasted
remains below 50\% while still allowing values of various sizes to be inserted.
In the following description, we will assume that values are of a fixed size.

@@

According to this description (which follows the paper by Bender et al) a window
is rebalanced every time an element is inserted. As an optimization, our
implementation does not always rebalance a window. When there is free space
between the successor and predecessor of the value to be inserted, the value
is simply inserted in the middle. In our experiments, this optimization
yielded a reduction in execution time.

@@

Finally, the Bender set uses a complete binary tree as an index data structure
to efficiently find the successor element for a given value. The tree is
stored in memory in van Emde Boas layout (as described by Prokop) to allow
efficient cache-oblivious traversal.

@@

Therefore, we decided to measure wall clock time. To deal with variations
across multiple executions, each experiment is run five times, and the median
value is used for further analysis.

@@

In the results below, all values are reported relative to the values obtained
using the mock set implementation, which means the overhead of the test
framework is removed from the results. Although the extra memory used by
the test framework is relatively small (less than 10\% in all cases) and
primarily caused by the queue data structure, the overhead in terms of
execution time was quite significant: close to 70\% on the worst case.
This does mean that the set data structure accounts for at least 30\%
of the execution time of the search algorithm in all cases (and much
more for the slower data structures).

\subsubsection{System Configuration}
The experiments where performed on a 64-bit Linux (kernel 2.6) system
with an Intel Xeon E5320 processor (1.86 GHz, 4~MB cache) and 8~GB of main
memory and 2~GB swap file. Although the system is a multiprocessor system
(and the E5320 is a dual-core processor) all code is single-threaded so
only a single core is used when executing tests.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l l r r }
\hline
\textbf{Model} & \textbf{Parameters} & \textbf{Iterations} & \textbf{Transitions} \\
\hline
Eratosthenes & $N=40$ &  1,019,960 &  4,923,218 \\
Leader2      & $N=5$  &  5,950,945 & 23,856,363 \\
Peterson\_N  & $N=4$  & 10,000,000 & 37,434,411 \\
\hline
\end{tabular}
\caption{Properties of test cases used}
\label{tab-cases}
\end{center}
\end{table}

@@

Figure \ref{fig-eratosthenes-hash_wctime} shows the execution times for the hash
table.
The hash table
with a small index (100,000 slots) performs well initially, but gets slower as
it becomes too full, at which point each slot in the hash table has to store a
long list of values that map to that slot.
As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
usage by the different hash tables is in the (fixed) size of the index.
We will use the hash table with an index size of 10,000,000 for further
experiments, as it seems to perform best even for a relatively small test case.

Figure \ref{fig-eratosthenes-btree_wctime} and Figure
\ref{fig-eratosthenes-btree_vss} show the execution time and memory usage
for the B-tree with page sizes ranging from 1~KB to 16~KB. The execution platform
maps memory in 4~KB pages, which would suggest that using pages less than 4KB
makes little sense. Indeed, the B-tree with 1~KB pages seems to perform worst,
but if we take the memory use into account, this is more likely due to the fact
that fewer items fit on a single page (which means a relatively large portion of
the page remains unused) than anyting else.
The 4~KB and 16~KB pages are relatively close (both in execution time and memory
usage); we select the 16~KB page size for further experiments because state sizes
are larger in other test cases, in which case the 4~KB page size may cause similar
problems as the 1~KB page size here.

@@

Now that we have established the parameters to use for the other test cases,
we can run final experiments on all three test cases.
The execution times for the these cases are presented in figures
\ref{fig-eratosthenes_wctime}, \ref{fig-leader2_wctime} and
\ref{fig-peterson_wctime}, while the corresponding memory usage is presented in
figures \ref{fig-eratosthenes_vss}, \ref{fig-leader2_vss} and \ref{fig-peterson_vss} respectively.

@@

It should be noted that the Bender set is the only data structure that shows
abrupt changes in both the execution time and memory usage graphs. These
jumps in the graph occur whenever the capacity of the Bender set is increased;
this causes the memory allocated for the set to be doubled and the array and
tree structures to be recreated, which is a relatively expensive operation.

@@

\subsection{Implementation Complexity}
Since all data structures used were implemented from scratch, our research
also yields some insight in the implementation complexity of the different
data structures.

@@

\section{Results}

@@
